---
title: "TimeSeries_FinalProject"
author: "vemani venkatesh"
date: "2024-11-22"
output:
  pdf_document: default
  html_document: default
---

```{r}
Gold_Prices <- read.csv("~/Downloads/Gold  Prices 2.csv")
View(Gold_Prices)
```

```{r}
length(unique(Gold_Prices$Date))

```

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

```

```{r}
# Check for missing values in the Date column
summary(Gold_Prices$Date)

# Generate a sequence of weekdays starting from 2019-11-11, excluding weekends
dates_seq <- seq(from = as.Date("2019-11-11"), by = "day", length.out = nrow(Gold_Prices))

# Remove weekends from the sequence
weekdays_only <- dates_seq[!weekdays(dates_seq) %in% c("Saturday", "Sunday")]

# Assign the weekday sequence to the Date column
Gold_Prices$Date <- weekdays_only[1:nrow(Gold_Prices)]

# Convert the Date column to POSIXct format
Gold_Prices$Date <- as.POSIXct(Gold_Prices$Date, format = "%Y-%m-%d", tz = "America/New_York")

```




```{r}

sum(is.na(Gold_Prices))

colSums(is.na(Gold_Prices))

```

```{r}
# Load necessary library
library(lubridate)

# Define the start date and calculate a range of weekdays
start_date <- as.Date("2023-04-22")  # Adjust based on your dataset's start date
end_date <- start_date + years(5) - 1  # Create a 5-year range
all_dates <- seq(from = start_date, to = end_date, by = "day")
weekdays_only <- all_dates[!weekdays(all_dates) %in% c("Saturday", "Sunday")]

# Check for NA values in the Date column
missing_count <- sum(is.na(Gold_Prices$Date))  # Count missing dates

# Fill NA values in the Date column with weekdays-only dates
Gold_Prices$Date[is.na(Gold_Prices$Date)] <- weekdays_only[1:missing_count]

# Print the updated dataset
print(Gold_Prices)


```
## Open price over time vs date year and close price over time vs date year.
```{r}

# Load necessary libraries
library(ggplot2)
library(gridExtra)

# Assuming your data frame is named df and has columns Open, Close, High, Low, and a Date column

# Create individual plots
p1 <- ggplot(Gold_Prices, aes(x = Date, y = Open)) +
  geom_line(color = "green") +
  labs(title = "Open Price Over Time", y = "Price") +
  theme_minimal() +
  theme(axis.title.x = element_blank(), plot.margin = margin(10, 10, 0, 10))

p2 <- ggplot(Gold_Prices, aes(x = Date, y = Close)) +
  geom_line(color = "blue") +
  labs(title = "Close Price Over Time", y = "Price") +
  theme_minimal() +
  theme(axis.title.x = element_blank(), plot.margin = margin(10, 10, 0, 10))

p3 <- ggplot(Gold_Prices, aes(x = Date, y = High)) +
  geom_line(color = "red") +
  labs(title = "High Price Over Time", y = "Price") +
  theme_minimal() +
  theme(axis.title.x = element_blank(), plot.margin = margin(10, 10, 0, 10))
p4 <- ggplot(Gold_Prices, aes(x = Date, y = Low)) +
  geom_line(color = "orange") +
  labs(title = "Low Price Over Time", x = "Date", y = "Price") +
  theme_minimal() +
  theme(plot.margin = margin(10, 10, 10, 25))

# Arrange the plots in a single figure
grid.arrange(p1, p2, ncol = 1)

```
# high price over time vs date year and low price over time vs date year.
```{r}
# Arrange the plots in a single figure
grid.arrange(p3, p4, ncol = 1)

```
## Trading Volume overtime vs Date Year
```{r}
# Load necessary libraries
library(ggplot2)
library(lubridate)

# Assuming your dataset is named Gold_Prices and it contains 'Date' and 'Volume' columns

# Ensure 'Date' column is in Date format
Gold_Prices$Date <- as.Date(Gold_Prices$Date)

# Plot the trading volume over time
ggplot(Gold_Prices, aes(x = Date, y = Volume)) +
  geom_line(color = "black") +           # Line chart for the volume
  theme_minimal() +                     # Use a clean theme
  labs(title = "Trading Volume Over Time",
       x = "Date", y = "Volume") +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year") +  # Format x-axis to show years
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") + # Add baseline
  theme(panel.grid.major = element_line(color = "grey", size = 0.5)) # Display grid

```


```{r}
# Load necessary libraries
library(ggplot2)

# Assuming Gold_Prices is your data frame with 'Date', 'Close', and 'Volume' columns
# Convert 'Date' to Date type if necessary
Gold_Prices$Date <- as.Date(Gold_Prices$Date)

# Create the plot
p <- ggplot(Gold_Prices, aes(x = Date)) +
  # Plot the closing price on the first y-axis
  geom_line(aes(y = Close), color = 'blue', linewidth = 1) +  
  ylab('Price') +
  ggtitle('Gold Prices Closing and Trading Volume Over Time') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  
  # Add a second y-axis for trading volume
  geom_bar(aes(y = Volume / 1e6), stat = 'identity', fill = 'red', alpha = 0.5) +  # Rescale Volume for better visibility
  scale_y_continuous(
    name = 'Price',
    sec.axis = sec_axis(~ . * 1e6, name = 'Volume')  # Scale back the volume on the secondary axis
  ) +
  theme(legend.position = 'top')

# Display the plot
print(p)

```
```{r}
# Load necessary libraries
library(ggplot2)

# Assuming Gold_Prices is your data frame with 'Date', 'Close', and 'Volume' columns
# Convert 'Date' to Date type if necessary
Gold_Prices$Date <- as.Date(Gold_Prices$Date)

# Create the plot
p <- ggplot(Gold_Prices, aes(x = Date)) +
  # Plot the closing price on the first y-axis
  geom_line(aes(y = Open), color = 'blue', linewidth = 1) +  
  ylab('Price') +
  ggtitle('Gold Prices Opening and Trading Volume Over Time') +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  
  # Add a second y-axis for trading volume
  geom_bar(aes(y = Volume / 1e6), stat = 'identity', fill = 'red', alpha = 0.5) +  # Rescale Volume for better visibility
  scale_y_continuous(
    name = 'Price',
    sec.axis = sec_axis(~ . * 1e6, name = 'Volume')  # Scale back the volume on the secondary axis
  ) +
  theme(legend.position = 'top')

# Display the plot
print(p)


```
# Data Preprocessing
```{r}
# Unique values in the 'Close' column
unique(Gold_Prices$Dividends)
unique(Gold_Prices$Stock.Splits)
unique(Gold_Prices$Capital.Gains)
```

# Dropped unecessary columns
```{r}

# Drop 'Dividends' and 'Stock.Splits' columns
Gold_Prices <- Gold_Prices[, !colnames(Gold_Prices) %in% c("Dividends", "Stock.Splits", "Capital.Gains")]
Gold_Prices

```

```{r}
str(Gold_Prices)

```

## Checking the Skewness
```{r}
# Load required library
library(ggplot2)
# Specify the feature columns
feature_columns <- c("Open", "High", "Low", "Volume", "Close")
# Plot histograms for features with normal distribution curve
for (feature in feature_columns) {
  print(ggplot(Gold_Prices, aes_string(x = feature)) + 
          geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7, aes(y = ..density..)) +  # Add density to histogram
          stat_function(fun = dnorm, args = list(mean = mean(Gold_Prices[[feature]], na.rm = TRUE), sd = sd(Gold_Prices[[feature]], na.rm = TRUE)), 
                        color = "blue", size = 1) +  # Add normal distribution curve
          ggtitle(paste("Distribution of", feature)) + 
          xlab(feature) + 
          ylab("Density"))
}


```


# Scatter plots to find the relationship between features and target column

```{r}

# Scatter plots of features vs. target
for (feature in feature_columns) {
  print(ggplot(Gold_Prices, aes_string(x = feature, y = "Close")) +
          geom_point(alpha = 0.6, color = "blue") +
          ggtitle(paste("Relationship between", feature, "and Close")) +
          xlab(feature) +
          ylab("Close"))
}


```
# applying log transformation
## for Target column "Close"
```{r}

# Apply log transformation to the 'Close' column
Gold_Prices$Log_Close <- log(Gold_Prices$Close)

# Preview the dataset to confirm the transformation
head(Gold_Prices)

```

```{r}

# Plot original vs. log-transformed Close prices
par(mfrow = c(1, 2))  # Arrange two plots in one row

# Original Close prices
plot(Gold_Prices$Date, Gold_Prices$Close, type = "l", col = "blue",
     main = "Original Close Prices", xlab = "Date", ylab = "Close")

# Log-transformed Close prices
plot(Gold_Prices$Date, Gold_Prices$Log_Close, type = "l", col = "red",
     main = "Log-Transformed Close Prices", xlab = "Date", ylab = "Log Close")


```
# applying log transformation
## for Feature "Volume"
```{r}

# Apply log transformation to the 'Close' column
Gold_Prices$Log_Volume <- log(Gold_Prices$Volume)

# Preview the dataset to confirm the transformation
head(Gold_Prices)

```
```{r}

# Plot original vs. log-transformed Close prices
par(mfrow = c(1, 2))  # Arrange two plots in one row

# Original Close prices
plot(Gold_Prices$Date, Gold_Prices$Volume, type = "l", col = "blue",
     main = "Original Volume", xlab = "Date", ylab = "Close")

# Log-transformed Close prices
plot(Gold_Prices$Date, Gold_Prices$Log_Volume, type = "l", col = "red",
     main = "Log-Transformed Volume ", xlab = "Date", ylab = "Log Close")


```

# checking the skewness after performing log transformation
```{r}
# Load required library
library(ggplot2)
# Define the feature columns manually
feature_columns <- c("Open","High", "Low","Log_Close","Log_Volume")
# Plot histograms for features with normal distribution curve
for (feature in feature_columns) {
  print(ggplot(Gold_Prices, aes_string(x = feature)) + 
          geom_histogram(bins = 30, fill = "skyblue", color = "black", alpha = 0.7, aes(y = ..density..)) +  # Add density to histogram
          stat_function(fun = dnorm, args = list(mean = mean(Gold_Prices[[feature]], na.rm = TRUE), sd = sd(Gold_Prices[[feature]], na.rm = TRUE)), 
                        color = "blue", size = 1) +  # Add normal distribution curve
          ggtitle(paste("Distribution of", feature)) + 
          xlab(feature) + 
          ylab("Density"))
}


```
# checking the linearity after log transformation.
```{r}

# Scatter plots of features vs. target
for (feature in feature_columns) {
  print(ggplot(Gold_Prices, aes_string(x = feature, y = "Close")) +
          geom_point(alpha = 0.6, color = "blue") +
          ggtitle(paste("Relationship between", feature, "and Close")) +
          xlab(feature) +
          ylab("Close"))
}


```

```{r}

# Drop 'Dividends' and 'Stock.Splits' columns
Gold_Prices <- Gold_Prices[, !colnames(Gold_Prices) %in% c("Close", "Volume")]
Gold_Prices

```
# ACF and PACF

```{r}
Log_Close_ts <- ts(Gold_Prices$Log_Close, start = c(2019, 11), frequency = 252)

```

# plotting the time series
```{r}
plot(Log_Close_ts, main = "Log_Close Time Series", ylab = "Log_Close", xlab = "Time")

```

# verifying the start date and end date and frequency of time series data 
```{r}
start(Log_Close_ts)  # Should return c(2019, 11)
end(Log_Close_ts)    # Should return the actual end date, e.g., c(2024, 5) or similar
frequency(Log_Close_ts)  # Should match the number of observations per year (252 or 365)

```

# performing ACF and PACF before applying differencing to the target
```{r}
# ACF plot
acf(Log_Close_ts, main = "Autocorrelation Function (ACF) of Log_Close")

# PACF plot
pacf(Log_Close_ts, main = "Partial Autocorrelation Function (PACF) of Log_Close")


```
# performing adf testing to check the stationarity 
```{r}
# Load library for ADF test
library(tseries)

# Perform ADF test
adf.test(Log_Close_ts)

```
# performing differencing
```{r}
Log_Close_diff <- diff(Log_Close_ts)

```

# performing adf testing again after performing the log differencing
```{r}
adf.test(Log_Close_diff)


```
# replacing the Log_Close with Log_Close_diff
```{r}
# Assuming your dataset is named 'data'
Gold_Prices <- Gold_Prices[-1258, ]  # Remove the first row

# Now replace the 'Log_Close' column with 'Log_Close_diff'
Gold_Prices$Log_Close <- Log_Close_diff

# Verify the change
Gold_Prices


```

# after applying differencing to the target

```{r}
Log_Close_ts1 <- ts(Gold_Prices$Log_Close, start = c(2019, 11), frequency = 252)
```


```{r}
# ACF plot
acf(Log_Close_ts1, main = "Autocorrelation Function (ACF) of Log_Close")

# PACF plot
pacf(Log_Close_ts1, main = "Partial Autocorrelation Function (PACF) of Log_Close")
```
# Splitting Data Train and Test Data ( 80:20)

```{r}
# Ensure Gold_Prices is your dataset and check it has data
if (nrow(Gold_Prices) > 0) {
  set.seed(123)  # Set seed for reproducibility
  
  # Calculate the size for the training set (80% of the data)
  train_size <- floor(0.8 * nrow(Gold_Prices))
  
  # Create the training and test sets based on the calculated index
  train_data <- Gold_Prices[1:train_size, ]
  test_data <- Gold_Prices[(train_size + 1):nrow(Gold_Prices), ]
  
  # Display the number of rows in each set
  cat("Training Set Size:", nrow(train_data), "\n")
  cat("Test Set Size:", nrow(test_data), "\n")
} else {
  cat("Dataset 'Gold_Prices' is empty. Please check your data.")
}
```

# Fitting ARIMA for the Training Data
```{r}
library(forecast)
# Ensure train_data is the column you want to model (e.g., 'Log_Close')
train_series <- train_data$Log_Close  # Replace 'Log_Close' with the desired column

# Convert to a time series object if it's not already
train_ts <- ts(train_series, frequency = 252)  # Use appropriate frequency for your data

# Fit ARIMA(1,1,1) model on the training data
model <- Arima(train_ts, order = c(1, 1, 1))

# Summarize the model
summary(model)


```

# checking for d =2
```{r}
# Ensure train_data is the column you want to model (e.g., 'Log_Close')
train_series <- train_data$Log_Close  # Replace 'Log_Close' with the desired column

# Convert to a time series object if it's not already
train_ts <- ts(train_series, frequency = 252)  # Use appropriate frequency for your data

# Fit ARIMA(1,1,1) model on the training data
model2 <- Arima(train_ts, order = c(1, 2, 1))

# Summarize the model
summary(model2)



```


```{r}
# Ensure the 'Date' column is in Date format
Gold_Prices$Date <- as.Date(Gold_Prices$Date)

# Creating a zoo object with 'Date' as the index for the 'Log_Close' series
library(zoo)
log_close_zoo <- zoo(Gold_Prices$Log_Close, order.by = Gold_Prices$Date)

# Fit the ARIMA model to the zoo object
library(forecast)
arima_model <- auto.arima(log_close_zoo)

# Extract residuals from the model
residuals <- residuals(arima_model)

# Convert residuals to a zoo object to preserve the correct dates
residuals_zoo <- zoo(residuals, order.by = index(log_close_zoo))

# Plot residuals with the correct date alignment
plot(residuals_zoo, main="ARIMA Residuals", xlab="Date", ylab="Residuals")


```

```{r}
library(zoo)

# Interpolate missing values
residuals <- na.approx(residuals)

```


```{r}
# Plot residuals
plot(residuals, main = "Residuals of ARIMA Model", ylab = "Residuals", xlab = "Time")

# Histogram
hist(residuals, main = "Histogram of Residuals", xlab = "Residuals", breaks = 20)


```
```{r}
# Ljung-Box test
ljung_box_test <- Box.test(residuals_zoo, lag = 4, type = "Ljung-Box")
print(ljung_box_test)


```


```{r}

# Ljung-Box test for autocorrelation in residuals
Box.test(residuals, lag = 2, type = "Ljung-Box")


```

```{r}
# Assuming 'model' is your fitted ARIMA model and 'test_data' is your test dataset

# Forecast the next values for the length of your test set
forecast_values <- forecast(model, h = nrow(test_data))

# Plot the forecasted values
plot(forecast_values, main = "ARIMA Forecast vs Actual Log_Close", 
     xlab = "Date", ylab = "Log_Close", col = "blue")

# Add the actual values (Log_Close) to the plot
lines(test_data$Date, test_data$Log_Close, col = "red", lwd = 2)

# Add a legend to differentiate actual vs forecasted
legend("topright", legend = c("Forecast", "Actual"), 
       col = c("blue", "red"), lty = 1, lwd = 2)

```


## Seasonal Decomposition of Time Series (STL) + Forecasting
```{r}
stl_decomp <- stl(train_ts, s.window = "periodic")
stl_forecast <- forecast(stl_decomp, h = length(test_data))
plot(stl_forecast)
summary(stl_forecast)
```

 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 
 